---
title: "Principal Component Analysis of the Hitters Dataset"
author: "Shradha Upadhyay"
date: "2026-08-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Principal Component Analysis of the Hitters Dataset

## Objective

This analysis performs Principal Component Analysis (PCA) on the predictor variables of the Hitters dataset. The goal is to examine the structure of the predictor variables, determine an appropriate number of principal components, and interpret the first two principal components.

Salary is the response variable and is excluded from the PCA.

## Dataset Overview

The `Hitters` dataset from the **ISLR** package contains Major League Baseball player statistics and salary information.

The original dataset contains **322 players and 20 variables**. After removing observations with missing salary values, **263 players** remain for the analysis.

The variables include:

- **Current-season performance:** `AtBat`, `Hits`, `HmRun`, `Runs`, `RBI`, and `Walks`
- **Career performance:** `CAtBat`, `CHits`, `CHmRun`, `CRuns`, `CRBI`, and `CWalks`
- **Experience:** `Years`
- **Fielding statistics:** `PutOuts`, `Assists`, and `Errors`
- **Categorical information:** `League`, `Division`, and `NewLeague`
- **Response variable:** `Salary`

Because the objective is to perform PCA on the predictors, `Salary` is excluded from the PCA analysis.
```{r}
# Load required libraries
library(ISLR)

# Load the Hitters dataset
data(Hitters)
```

```{r}
# Check the dimensions of the dataset
dim(Hitters)
```
```{r}
# View variable names
names(Hitters)
```
```{r}
# View the first few observations
head(Hitters)
```
## Data Preparation
```{r}
# Check the number of missing values in each variable
colSums(is.na(Hitters))
```
```{r}
# Remove observations with missing values
Hitters_clean <- na.omit(Hitters)

# Check dimensions after removing missing values
dim(Hitters_clean)
```

## Predictor Variables
```{r}
# Check the structure and data types of the variables
str(Hitters_clean)
```
## (a)  Should the Predictors Be Standardized?

Yes. Standardization is appropriate because the Hitters predictors are measured on very different scales. For example, career statistics such as `CAtBat` can be in the thousands, whereas `Years` is much smaller.

Without standardization, variables with larger variances could dominate the principal components. Standardizing the predictors to mean 0 and standard deviation 1 allows PCA to compare variables on a common scale.
## (b) Principal Component Analysis

For the PCA, `Salary` is excluded because it is the response variable. The analysis is performed using only the predictor variables.
```{r}
# Remove Salary and keep only the predictor variables
X <- Hitters_clean[, names(Hitters_clean) != "Salary"]

# Check dimensions
dim(X)
```

```{r}
# Identify categorical predictor variables
names(X)[sapply(X, is.factor)]
```
```{r}
# Convert categorical predictors to dummy variables
X_numeric <- model.matrix(~ ., data = X)[, -1]

# Check dimensions
dim(X_numeric)
```

### Standardize the Predictor Variables
Because the predictors are measured on different scales, the variables are standardized to have mean 0 and standard deviation 1 before performing PCA.
```{r}
# Standardize the predictor variables
X_scaled <- scale(X_numeric)

# Check the dimensions
dim(X_scaled)
# Check the mean and standard deviation of the standardized predictors
round(colMeans(X_scaled), 3)
# Check standard deviations
round(apply(X_scaled, 2, sd), 3)
```
### Perform Principal Component Analysis

PCA is performed on the standardized predictor variables to obtain a smaller set of uncorrelated principal components that summarize the variation in the original predictors.
```{r}
# Perform PCA on the standardized predictors
pca_result <- prcomp(X_scaled)

# View PCA summary
summary(pca_result)
```
### Scree Plot

A scree plot is used to examine how much variance is explained by each principal component and to help determine an appropriate number of components to retain.
```{r}
# Calculate proportion of variance explained by each PC
variance_explained <- pca_result$sdev^2 / sum(pca_result$sdev^2)

# Open PNG file
png(
  "Figures/variance_explained.png",
  width = 800,
  height = 600
)

# Plot proportion of variance explained
plot(
  variance_explained,
  type = "b",
  pch = 19,
  xlab = "Principal Component",
  ylab = "Proportion of Variance Explained",
  main = "Variance Explained by Principal Components"
)

# Save and close the file
dev.off()
```
### Interpretation

The first principal component explains approximately **38.31%** of the total variance, while the second explains approximately **21.84%**. Together, the first two PCs explain about **60.16%** of the total variance.

The scree plot shows a substantial decrease in explained variance over the first few components and begins to flatten after approximately PC3–PC4. The first four principal components together explain approximately **79.03%** of the total variance.

Based on the scree plot and cumulative variance explained, retaining approximately **4 principal components** provides a reasonable lower-dimensional representation of the predictor data.

### Variance Explained by Principal Components
```{r}
# Calculate cumulative variance explained
cumulative_variance <- cumsum(variance_explained)

# Create PCA variance table
pca_variance_table <- data.frame(
  PC = paste0("PC", 1:length(variance_explained)),
  Variance_Explained = round(variance_explained, 4),
  Cumulative_Variance = round(cumulative_variance, 4)
)

# Display the table
pca_variance_table
# Display the first 10 principal components
head(pca_variance_table, 10)

```
## (c) Interpretation of the First Two Principal Components

The first two principal components are examined using correlations with the standardized quantitative predictor variables, PC scores, loadings, and a biplot.
```{r}
# Identify quantitative predictor variables
quantitative_vars <- names(X)[sapply(X, is.numeric)]

# View quantitative variable names
quantitative_vars

```
```{r}
# Keep only the quantitative predictor variables
X_quantitative <- X[, quantitative_vars]

# Check dimensions
dim(X_quantitative)
```


```{r}
# Standardize the quantitative predictor variables
X_quant_scaled <- scale(X_quantitative)

# Check dimensions
dim(X_quant_scaled)
```

```{r}
# Extract scores for the first two principal components
PC_scores <- pca_result$x[, 1:2]

# Check dimensions
dim(PC_scores)
```

```{r}
# Calculate correlations between quantitative variables and PC1 and PC2
pc_correlations <- cor(X_quant_scaled, PC_scores)

# View correlations
round(pc_correlations, 3)
```

### Correlation Interpretation

PC1 is strongly positively correlated with career statistics such as `CRBI`, `CRuns`, `CAtBat`, `CHits`, `CHmRun`, and `CWalks`, as well as `Years`. Therefore, PC1 primarily represents **career experience and accumulated career performance**.

PC2 is strongly positively correlated with current-season statistics such as `AtBat`, `Hits`, `Runs`, and `RBI`, while several career statistics have negative correlations with PC2. Therefore, PC2 appears to represent **current-season offensive performance relative to accumulated career performance**.

### Correlations of Quantitative Variables with PC1 and PC2

The correlations between the standardized quantitative predictors and the first two principal components are visualized below.

```{r}
# Save correlation plot
png(
  "Figures/pc1_pc2_correlations.png",
  width = 800,
  height = 600
)

plot(
  pc_correlations[, 1],
  pc_correlations[, 2],
  xlab = "Correlation with PC1",
  ylab = "Correlation with PC2",
  main = "Correlations of Variables with PC1 and PC2",
  pch = 19
)

text(
  pc_correlations[, 1],
  pc_correlations[, 2],
  labels = rownames(pc_correlations),
  pos = 3,
  cex = 0.8
)

abline(h = 0, v = 0, lty = 2)

dev.off()
```
### PC1 and PC2 Loadings
```{r}
# Extract loadings for PC1 and PC2
loadings_pc12 <- pca_result$rotation[, 1:2]

# Display the loadings
round(loadings_pc12, 3)
```

### Biplot of PC1 and PC2

The biplot displays the player scores on the first two principal components together with the variable loadings.
```{r}
# Save PCA biplot
png(
  "Figures/pca_biplot_pc1_pc2.png",
  width = 900,
  height = 700
)

biplot(
  pca_result,
  choices = c(1, 2),
  xlabs = rep(".", nrow(pca_result$x)),
  main = "PCA Biplot: PC1 and PC2",
  cex = 0.7
)

dev.off()
```

### Biplot Interpretation

The biplot displays the scores of the 263 players on PC1 and PC2 together with the predictor loadings.

Variables related to career performance, such as `CAtBat`, `CHits`, `CRuns`, `CRBI`, `CWalks`, and `Years`, load strongly in the positive PC1 direction and generally in the negative PC2 direction. In contrast, current-season variables such as `AtBat`, `Hits`, `Runs`, and `RBI` load positively on both PC1 and PC2.

This supports the earlier correlation analysis: **PC1 primarily represents overall career experience and accumulated performance**, while **PC2 helps distinguish current-season offensive performance from accumulated career performance**.

Players located in the direction of a particular group of variables tend to have relatively larger values for those variables.

## Conclusion

PCA was applied to the standardized predictor variables in the Hitters dataset to reduce dimensionality and summarize the major patterns in the data.

The first principal component explained approximately **38.31%** of the total variance, and the first two components together explained about **60.16%**. Based on the scree plot and cumulative variance, retaining approximately **4 principal components** provides a reasonable lower-dimensional representation, capturing about **79.03%** of the total variance.

The interpretation of PC1 and PC2 showed that **PC1 primarily reflects career experience and accumulated performance**, while **PC2 is more strongly associated with current-season offensive performance relative to career totals**.

Overall, the analysis demonstrates how PCA can summarize a large set of correlated baseball statistics using a smaller number of meaningful principal components.

