---
title: "Clustering Analysis of the Hitters Dataset"
author: "Shradha Upadhyay"
date: "2026-08-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Clustering Analysis of the Hitters Dataset

## Objective

This analysis applies unsupervised clustering methods to the Hitters dataset to identify groups of players with similar characteristics.

Two clustering approaches are considered:

- Hierarchical clustering using complete linkage
- K-means clustering with K = 2

The resulting clusters are compared and interpreted using player characteristics and mean salary.
## Dataset Overview

The `Hitters` dataset from the **ISLR** package contains Major League Baseball player statistics and salary information.

After removing observations with missing values, the dataset contains **263 players and 20 variables**. `Salary` is used to summarize the resulting clusters but is not used to form the clusters.

## Load and Prepare the Data
```{r}
# Load the Hitters dataset
library(ISLR)

data(Hitters)

# Remove observations with missing values
Hitters_clean <- na.omit(Hitters)

# Check dimensions
dim(Hitters_clean)
```
## (a) Should the Variables Be Standardized?

Yes. Standardization is important before clustering because the Hitters variables are measured on very different scales. For example, career statistics such as `CAtBat` can have values in the thousands, while variables such as `Years` have much smaller values.

Because distance-based clustering is affected by the scale of the variables, predictors with larger numerical ranges could dominate the distance calculations. Standardizing the variables to mean 0 and standard deviation 1 allows each predictor to contribute on a comparable scale.
## (b) Choice of Distance Measure

A **metric-based distance**, specifically Euclidean distance, is appropriate for clustering the players. The goal is to group players who are similar in their standardized baseball statistics.
Because the variables are standardized before clustering, Euclidean distance can compare players across the different predictor variables without variables with larger scales dominating the distance calculation.
## (c) Hierarchical Clustering

The players are clustered using the predictor variables only. `Salary` is excluded from the clustering variables and will later be used to compare the resulting clusters.
```{r}
# Remove Salary and keep only predictor variables
X <- Hitters_clean[, names(Hitters_clean) != "Salary"]

# Check dimensions
dim(X)
```
```{r}
# Convert categorical predictors to dummy variables
X_numeric <- model.matrix(~ ., data = X)[, -1]

# Check dimensions
dim(X_numeric)
```
### Standardize the Predictor Variables

The predictor variables are standardized before clustering so that variables measured on larger scales do not dominate the Euclidean distance calculation.
```{r}
# Standardize the predictor variables
X_scaled <- scale(X_numeric)

# Check dimensions
dim(X_scaled)
```
### Calculate Euclidean Distance

Euclidean distance is calculated between players using their standardized predictor values. Players with smaller distances have more similar baseball characteristics.
```{r}
# Calculate Euclidean distance between players
distance_matrix <- dist(X_scaled, method = "euclidean")

# View the first few distances
head(distance_matrix)
```
### Complete-Linkage Hierarchical Clustering

Hierarchical clustering is performed using complete linkage and the Euclidean distances calculated from the standardized predictor variables.
```{r}
# Perform hierarchical clustering using complete linkage
hc_complete <- hclust(
  distance_matrix,
  method = "complete"
)

# View clustering result
hc_complete
```
### Hierarchical Clustering Dendrogram

The dendrogram shows how individual players and clusters of players are progressively joined based on their similarity.

```{r}
# Save hierarchical clustering dendrogram
png(
  "Figures/clustering_hierarchical_dendrogram.png",
  width = 900,
  height = 650
)

# Plot dendrogram
plot(
  hc_complete,
  labels = FALSE,
  main = "Complete-Linkage Hierarchical Clustering",
  xlab = "Players",
  ylab = "Height"
)

# Highlight the two clusters
rect.hclust(
  hc_complete,
  k = 2,
  border = 2:3
)

# Close and save the figure
dev.off()
```
### Cut the Dendrogram into Two Clusters

The hierarchical clustering tree is cut to produce two distinct groups of players.
```{r}
# Cut the dendrogram into 2 clusters
hc_clusters <- cutree(hc_complete, k = 2)

# Check the number of players in each cluster
table(hc_clusters)
```
```{r}

# Save hierarchical clustering dendrogram
png(
  "Figures/clustering_hierarchical_dendrogram.png",
  width = 900,
  height = 650
)

# Plot dendrogram
plot(
  hc_complete,
  labels = FALSE,
  main = "Complete-Linkage Hierarchical Clustering",
  xlab = "Players",
  ylab = "Height"
)

# Highlight the two clusters
rect.hclust(
  hc_complete,
  k = 2,
  border = 2:3
)

# Save and close the figure
dev.off()
```
### Cluster-Specific Means

To interpret the two clusters, the mean values of the quantitative player characteristics are compared across the hierarchical clusters.
```{r}
# Select quantitative predictor variables
quantitative_vars <- names(X)[sapply(X, is.numeric)]

X_quantitative <- X[, quantitative_vars]

# Check dimensions
dim(X_quantitative)
```
```{r}
# Add hierarchical cluster labels
Hitters_clean$HC_Cluster <- factor(hc_clusters)

# Calculate cluster-specific means
hc_means <- aggregate(
  X_quantitative,
  by = list(Cluster = Hitters_clean$HC_Cluster),
  FUN = mean
)

# Round only numeric columns
hc_means[, -1] <- round(hc_means[, -1], 2)

# Display cluster means
hc_means
```
### Mean Salary by Cluster

The mean salary of players in each cluster is calculated to examine whether the groups identified by hierarchical clustering also differ in salary.

```{r}
# Calculate mean salary for each hierarchical cluster
hc_salary <- aggregate(
  Salary ~ HC_Cluster,
  data = Hitters_clean,
  FUN = mean
)

# Round mean salaries
hc_salary$Salary <- round(hc_salary$Salary, 2)

# Display results
hc_salary
```

### Hierarchical Clustering Interpretation

Complete-linkage hierarchical clustering separated the 263 players into two clusters containing **233** and **30** players.

The current-season statistics, such as `AtBat`, `Hits`, and `Runs`, were relatively similar between the two clusters. However, Cluster 2 showed substantially higher career statistics and experience. For example, the average number of `Years` was **15.83** in Cluster 2 compared with **6.21** in Cluster 1, while average career at-bats (`CAtBat`) were **7,233.50** compared with **2,068.36**.

The mean salary was also substantially higher in Cluster 2 (**970.68**) than in Cluster 1 (**479.95**).

Overall, the hierarchical clustering primarily separates a smaller group of **highly experienced players with large accumulated career statistics and higher salaries** from a larger group of players with less career experience and lower average salaries.

## (d) K-Means Clustering with K = 2

K-means clustering is performed on the standardized predictor variables using two clusters. The resulting clusters are summarized using the player characteristics and mean salary.
```{r}
# Set seed for reproducibility
set.seed(123)

# Perform K-means clustering with K = 2
kmeans_result <- kmeans(
  X_scaled,
  centers = 2,
  nstart = 25
)

# Number of players in each cluster
table(kmeans_result$cluster)
```
```{r}
# Add K-means cluster labels
Hitters_clean$KMeans_Cluster <- factor(kmeans_result$cluster)

# Calculate cluster-specific means for quantitative predictors
kmeans_means <- aggregate(
  X_quantitative,
  by = list(Cluster = Hitters_clean$KMeans_Cluster),
  FUN = mean
)

# Round numeric columns
kmeans_means[, -1] <- round(kmeans_means[, -1], 2)

# Display cluster means
kmeans_means
```

```{r}
# Calculate mean salary for each K-means cluster
kmeans_salary <- aggregate(
  Salary ~ KMeans_Cluster,
  data = Hitters_clean,
  FUN = mean
)

# Round mean salary
kmeans_salary$Salary <- round(kmeans_salary$Salary, 2)

# Display results
kmeans_salary
```
### K-Means Clustering Interpretation

K-means clustering with K = 2 divided the 263 players into **189 players in Cluster 1** and **74 players in Cluster 2**.

Cluster 2 had higher average current-season statistics, greater career experience, and substantially larger career totals than Cluster 1. For example, Cluster 2 had higher average `Hits`, `HmRun`, `RBI`, `Years`, `CAtBat`, and `CHits`.

The mean salary was also substantially different between the groups. Cluster 1 had a mean salary of **368.55**, whereas Cluster 2 had a mean salary of **963.40**.

Overall, K-means separated the players into a group with generally lower performance and experience and a group of more experienced, higher-performing players with substantially higher average salaries.
## (e) Comparison of Hierarchical and K-Means Clustering

The cluster assignments from hierarchical clustering and K-means clustering are compared to determine how similarly the two methods grouped the players.

```{r}
# Compare hierarchical and K-means cluster assignments
cluster_comparison <- table(
  Hierarchical = hc_clusters,
  KMeans = kmeans_result$cluster
)

cluster_comparison

```
### Comparison of Cluster Assignments

The two clustering methods showed substantial agreement in their cluster assignments.

All **189 players** assigned to K-means Cluster 1 were also assigned to hierarchical Cluster 1. In addition, all **30 players** in hierarchical Cluster 2 were also assigned to K-means Cluster 2.

However, **44 players** assigned to hierarchical Cluster 1 were placed in K-means Cluster 2. Therefore, K-means produced a larger second cluster than hierarchical clustering.

Overall, both methods identified a common group of 30 players, while K-means classified an additional 44 players into the higher-performance and more experienced group.
## Conclusion

Both hierarchical clustering and K-means clustering identified meaningful groups of players based on their standardized baseball statistics.

Complete-linkage hierarchical clustering produced two clusters containing **233 and 30 players**. The smaller cluster consisted primarily of more experienced players with substantially larger career statistics and a higher mean salary (**970.68 vs. 479.95**).

K-means clustering with K = 2 produced clusters containing **189 and 74 players**. The second cluster also represented more experienced, higher-performing players and had a substantially higher mean salary (**963.40 vs. 368.55**).

The two methods agreed on a core group of **30 players**, while K-means assigned an additional **44 players** to its higher-performance cluster.

Overall, both clustering approaches revealed a relationship between player experience, career performance, and salary, although the exact cluster assignments differed between the two methods.